## **Data Visualization 02 - Term Project**
### by Fatima Arshad

Formula One (also known as Formula 1 or F1) is the highest class of international racing for open-wheel single-seater formula racing cars sanctioned by the Fédération Internationale de l'Automobile (FIA). The World Drivers' Championship, which became the FIA Formula One World Championship in 1981, has been one of the premier forms of racing around the world since its inaugural season in 1950. Formula One cars are the fastest regulated road-course racing cars in the world, owing to very high cornering speeds achieved through the generation of large amounts of aerodynamic downforce. A Formula One season consists of a series of races, known as Grands Prix, which take place worldwide on both purpose-built circuits and closed public roads. Formula 1 recently completed its 70 years anniversary. Lewis Hamilton clinched his 7th world title equaling Michael Schumachers record. The purpose of this study is to explore the best F1 drivers over the past 70 years.


## **Dataset**

The data was accessed from the its official website of [Formula One](https://www.formula1.com/en/results.html/2020/races.html) and is uploaded on my [github] (https://raw.githubusercontent.com/fatimamalikk/data-visualization-2/master/project_data). The website has a results page that lists the results for all seasons since 1950. Success of a driver was measured using the number of races won.

```{r, include=FALSE, message=FALSE, warning=FALSE}
# Set Cran mirror to avoid error
local({r <- getOption("repos")
       r["CRAN"] <- "http://cran.r-project.org" 
       options(repos=r)
})

rm(list = ls())

# Load libraries
library(tidyverse)
library(data.table)
library(rvest)
library(gganimate)
library(stringr)
#install.packages("gifski")
library(gifski)
library(knitr)
```

The following dataframe shows the race results for all the seasons in range 1050- 2020:

```{r, echo = FALSE, message= FALSE,warning = FALSE}
database <- read_csv('https://raw.githubusercontent.com/fatimamalikk/data-visualization-2/master/project_data/f1_results.csv')
table_format <- tail(database)
kable(table_format)

```

## Data Visualization

### Top 5 most no of wins by F1 drivers (1950-2020)

The bar graph below visualizes the most successful drivers of seasons 1050-2020:

```{r, fig.width = 8, fig.height = 5 , echo = FALSE, message= FALSE,warning = FALSE}

database %>% 
  group_by(winner) %>%
  summarize(total_wins = n()) %>%
  arrange(desc(total_wins)) %>%
  head(5) %>%
  ggplot(aes(x = reorder(winner, total_wins), y = total_wins, fill = winner,label = total_wins)) + geom_col() + coord_flip() +geom_text(hjust = -0.3) + theme_classic() + theme(legend.position = "none",plot.title=element_text(hjust = 0.5,face="bold")) +
  labs( x = "F1 Race Winners", y = "Total no of race wins", title = "Top 5 most no of wins by F1 drivers (2010-2020)")
  
```

As evident from the figure above, Lewis Hamilton has been the most successful F1 driver of all time with 95 wins followed by Michael Schumacher with 91 wins and Sebastian Vettel on third position with 53 wins.

### Top 5 most successful drivers winning progression over their career

We previously found out the top 5 most successful drivers for the tenure 1050-2020. Next, we will take a deeper dive into how the top 5 drivers secured wins over the their racing careers:

```{r, fig.width = 11, fig.height=5.5, echo = FALSE, message= FALSE,warning = FALSE}

# Extract years from date column
database$year <-str_sub(database$race_date, start= -4)

#Filter database for top 5 most successful drivers
drivers <- c("Lewis Hamilton","Michael Schumacher", "Sebastian Vettel", "Alain Prost", "Ayrton Senna")
top_5_stats<- database %>% filter(winner %in% drivers)

# Convert year column to int
top_5_stats <- top_5_stats %>% mutate(year = as.numeric(year))

# Aggregate no of wins for each year for the 5 drivers
top_5_per_year<- top_5_stats %>% 
  group_by(year,winner) %>%
  count(winner)

# Calculate the cumulative sum of the wins per year for each driver
top_5_per_year_cs <- top_5_per_year %>% group_by(winner) %>% arrange(winner,year) %>% mutate(cs = cumsum(n))

# Visualize the no of wins for each driver over their racing career  
top_5_per_year_cs %>% ggplot(aes(x = year, y = cs, fill = winner,group = winner)) + geom_line(aes(colour=winner),size = 2) +
  labs( x = "Year",
        y = "Races Won",
        title = "Race wins progression of Top 5 most successful drivers over their careers") + 
  scale_y_continuous(breaks = seq(0,100, by=20)) + 
  scale_x_continuous(breaks = seq(1978,2020, by=4)) + 
  theme_classic() +
  theme(legend.position = "none",
        axis.title=element_text(size=14,face="bold"),
        plot.title=element_text(size = 14,face = "bold", hjust = 0.4),
        axis.text = element_text(size = 14,vjust = 2.5,hjust = 0.7)) +
    annotate("text", x = 1994, y = 57, label = "Alain Prost", fontface = 2) +
    annotate("text", x = 1995, y = 45, label = "Ayrton Senna", fontface = 2) +
    annotate("text", x = 2008, y = 96, label = "Michael Schumacher", fontface = 2) +
    annotate("text", x = 2018, y = 99, label = "Lewis Hamilton", fontface = 2) +
    annotate("text", x = 2019, y = 40, label = "Sebastian Vettel", fontface = 2)
  
  
```

Lewis Hamilton career wins are the most impressive from 2014 onward with a steep addition in race wins. Michael Schumachers race wins rate is nearly identical to Lewis Hamilton but with an occasional slow down in some periods. Both had careers ranging approximately 12 years over which they gathered over 90 wins each. Sebestian Vettel had a better start compared to his rival Lewis Hamilton bagging a lot wins and consequently championships from 2010-2013. From 2014 onwards with his move to Ferrari, the rate of race wins decreased significantly. Alain Prost gathered 51 race wins over his 11 year racing period at an average of 5 race wins a season. Ayrton Senna had a slow start with 6 wins only in his first 4 years but then displayed dominant performance over the next 6 years until his death during a race crash.

### Visualize F1 racing tracks of top 10 countries

The bar graph below visualizes the most successful drivers of seasons 1050-2020:

```{r, fig.width = 11, fig.height=5.5, echo = FALSE, message= FALSE,warning = FALSE}
df <- as.data.table(database)
df_countries <- df[,
                   .(mean_lap = mean(as.numeric(laps), na.rm=TRUE)),
                   by = .(race_name) ]
#df_countries[order(df_countries$mean_lap)]  %>% top_n(10)

df_top_countries <- df_countries[order(df_countries$mean_lap),] %>% top_n(10)


ggplot(data=df_top_countries, aes(x=race_name, y=mean_lap,group=1)) +
  geom_line(color="salmon", size=2) +
  geom_point(color="chocolate", size=3.5) +
  ggtitle("Scatterplot of top 10 F1 race tracks") +
  labs(x="", y="Average value of laps\nper race track") +
  theme(axis.title.y = element_text(size=7)) +
  theme(axis.text = element_text(size=6)) +
  theme(plot.title = element_text(size=10, hjust=0)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))
  #theme_bw()


```
The figure above shows the average laps for the top 10 tracks completed over an average of one hour and thirty minutes. The number of laps in a race is obtained by dividing 305 by the length of a lap, which differs from track to track. The duration of the race cannot be more than 2 hours. If the allocated time of 2 hours is exceeded, the race is considered to be finished at the end of the ongoing lap. The line graph above shows that drivers could finish the greatest number of laps i.e. 194 on the racing track situated in Indianapolis, followed by Sakhir with 84 laps. 


### Visualize top ten winning teams in the history of formula1 race

The boxplot below visualizes the top ten winning teams in the history of formula 1 race with respect to lap time. Upon hover over the points a tooltip appears with the actual number of laps completed by the team.

```{r, fig.width = 11, fig.height=5.5, echo = FALSE, message= FALSE,warning = FALSE}
#df <- as.data.table(database)

top_teams_list <- list('Ferrari', 'Mercedes', 'McLaren Mercedes','Williams Renault', 'Lotus Ford', 'McLaren Honda', 'Red Bull Racing Renault', 'Renault', 'McLaren Ford', 'McLaren TAG')

df_teamlap <- df[,
              .(laps),
              by = .(team) ] %>% filter(team == top_teams_list)

df_teamlap <- df_teamlap[order(df_teamlap$laps),]


df_teamlap$team <- as.factor(df_teamlap$team)


#########################
# boxplot of strength within categories
library(ggiraph)
winning_teams <-
    ggplot(df_teamlap, aes(x = team, y = as.numeric(laps))) + 
    geom_boxplot_interactive(aes(fill = team), outlier.shape = NA) + 
    geom_jitter_interactive(aes(tooltip = laps, data_id = team, color = team), width = 0.55, height = 1, alpha = 0.25) + # use this layer for points with tooltips
    labs(x = '', y = 'Laps', title = 'Top ten winning teams ') +
    theme_light() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          legend.position = 'none')

# show the boxplot
girafe(ggobj = winning_teams,
       options = list(opts_hover_inv(css = "opacity:0.1;")))


```
The boxplot above illustrates top ten teams securing the highest number of wins in the history of formula1 racing. It also shows the laps completed by each team in each competition. The data points jitter around Ferrari in the figure confirms that the most wins were secured by ferrari team and the least wins amongst the ten secured by Mclaren TAG team. Moreover, it is visible that Lotus Ford and McLaren Ford have a wide range of number of laps completed in their entire racing career. 

### **Racing Bar Chart Visualization for Most Wins by F1 Drivers (2015-2020)**

Using the subset of results for 2015-2020 from our results dataframe to visualize it using a racing bar chart. It will show how drivers rankings for most wins have changed over past 5 years.

```{r, echo = FALSE, message= FALSE,warning = FALSE}

# Import a table of cumulative race wins for the past 5 years`

driversTable <- read_csv("https://raw.githubusercontent.com/fatimamalikk/data-visualization-2/master/project_data/cumulative_results.csv")

# Prepare data for ranking
win_ranking <- driversTable %>%
  group_by(year) %>%
  # The * 1 makes it possible to have non-integer ranks while sliding
  mutate(rank = min_rank(-value) * 1,
         Value_rel = value/value[rank==1],
         Value_lbl = paste0(" ",value)) %>%
  filter(rank <=10) %>%
  arrange(value,winner) %>%
  ungroup() %>%
  group_by(year) %>%
  arrange(rank, winner) %>%
  mutate(rank = seq(1, n())) %>%
  ungroup() 

# Build individual plots for each year
plots_year <- ggplot(win_ranking, aes(rank, group = winner, 
                     fill = as.factor(winner), color = as.factor(winner))) +
  geom_tile(aes(y = value/2,
                height = value,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(winner, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=value,label = Value_lbl, hjust=0),size = 7) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  
  labs(title = "Total no of wins by F1 Drivers from 2015 - {closest_state}", x = "", y = "Winners",
       caption = "Most No of wins by F1 Drivers 2015-2020") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, vjust = 1.5, size = 22),
        axis.ticks.y = element_blank(),  # These relate to the axes post-flip
        axis.text.y  = element_blank(),  # These relate to the axes post-flip
        plot.margin = margin(1,1,1,4, "cm")) +
  
  transition_states(year, transition_length = 3, state_length = 0) +
  ease_aes('cubic-in-out')
      

# Save as preferred rendered format
# It will take a minute or so to create GIF
gif <- animate(plots_year, 200, fps = 10, duration = 9, width = 800, height = 600, renderer = gifski_renderer("gganim.gif"))

```

```{r, echo = FALSE, message= FALSE,warning = FALSE}
knitr::include_graphics(gif)
```

## **Summary**

We visualized the results via various graphs and found out that Lewis Hamilton is the most successful F1 driver in terms of race wins followed closely by Michael Schumacher. The top 5 most successful drivers had less wins in initial 2-3 years of their racing career but went on to dominate for the next 5-6 years. For race results since 2015, again Lewis Hamilton has showed his dominant form and has the highest number of wins followed by Nico Rosberg who although retired after 2016 season still ranks in 2nd place. Lastly, team Ferrari was the unprecedented winner in the history of formula1 racing.